\documentclass[10pt,journal,compsoc]{IEEEtran}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage{cite}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    pdftitle={Operationalizing Semantic Data Engineering for Algorithmic Trade Execution},
    pdfpagemode=FullScreen,
}

\begin{document}

\title{Operationalizing Semantic Data Engineering for Algorithmic Trade Execution}

\author{Yufeng Chen, Marc Delgado, Manuel Rucker\\
\small{Department of Advanced Databases (BDA-GIA), Universitat Polit\`ecnica de Catalunya (UPC)}\\
\small{Date: May 23, 2026}}

\markboth{Semantic Data Management for AI, Project 2, May 2026}{}

\IEEEtitleabstractindextext{%
\begin{abstract}
Traditional quantitative finance pipelines rely on isolated tabular datasets and short-horizon binary classifiers, which fail to capture the relational structure of corporate markets and suffer from high transaction costs and silent failure modes when deployed. This paper presents an end-to-end operational semantic trading infrastructure. We construct a multi-layered financial and macroeconomic Knowledge Graph (KG) in our Exploitation Zone, linking corporate structures, sectors, and World Bank macroeconomic indicators via RDFLib. On top of this data fabric we deploy two analytical pipelines: a multi-graph SPARQL querying engine and a graph-neural-network pipeline in which RotatE structural embeddings are PCA-compressed and merged with technical and macroeconomic indicators to \emph{regress the 30-day cross-sectional rank} of each stock. The pivot to a relative-ordering target unlocks the structural KG signal: across an expanding-window walk-forward CV with $K{=}5$ folds and a 30-day embargo, the tabular-only soft-vote ensemble reaches a fold-mean Spearman information coefficient (IC) of only $0.040 \pm 0.074$, KG-only features alone climb to $0.098 \pm 0.070$ (XGBoost), and the combined set lands at $0.092 \pm 0.056$ with a decile long-short spread of $+7.54\% \pm 3.99\%$ per month. The semantic ranker drives an institutional-grade operational layer: a 2-state Gaussian Hidden Markov Model gates regime exposure, a state-space Kalman Beta Filter sizes systematic risk, and a differential portfolio optimiser executes only the weight deltas at a strict 10 basis points transaction-cost charge. The live agent is hardened with three production controls absent from typical academic prototypes: a reproducibility hash that stamps every order with its model fingerprint and feature snapshot, a peak-to-trough drawdown circuit breaker, and a three-bucket HMM/ML/Kalman performance attribution log.
\end{abstract}

\begin{IEEEkeywords}
Semantic Data Engineering, Knowledge Graphs, Relational Graph Embeddings, Hidden Markov Models, Kalman Filters, Portfolio Rebalancing, Algorithmic Trading.
\end{IEEEkeywords}}

\maketitle
\IEEEdisplaynontitleabstractindextext
\IEEEpeerreviewmaketitle

\section{Introduction}
\IEEEPARstart{I}{n} quantitative asset management, predicting asset price movements is only half the battle. A highly accurate machine learning model is of zero economic value if it cannot be operationalized within a robust data engineering infrastructure. Data pipelines in the real world must ingest heterogeneous data, homogenize it semantically, manage systemic risk dynamically, and execute trades while minimizing transaction costs, spreads, and market slippage.

The academic objective of this project is to extend a classical tabular machine learning pipeline (developed in Phase 1) into an end-to-end semantic data science pipeline (Phase 2). P1 established a structured data lake (Landing, Formatted, and Trusted zones) to clean historical NASDAQ tick prices and train baseline Multi-Layer Perceptrons (MLP) and Random Forests. P2 transforms this architecture by embedding a **Knowledge Graph (KG)** into the Exploitation Zone. 

However, we recognized that simply feeding static relational embeddings into a classifier is insufficient to add real-world value. Therefore, we went far beyond the standard guidelines. We developed a complete, production-grade **Algorithmic Trading Bot CLI runner** that operates under a strict **10 basis points (10 bps)** transaction fee model. It dynamically adjusts market exposures using a **Gaussian Hidden Markov Model (HMM)** and scales short exposures using a time-varying **Kalman Beta Filter**—creating an institutional-grade, closed-loop trading infrastructure.

\section{The Semantic Data Engineering Pipeline}
The system's backbone is a structured, semantic DataOps pipeline. The data flows sequentially across four defined zones:

\subsection{Landing, Formatted, and Trusted Zones}
Raw data is ingested via Python collectors from Yahoo Finance, Nasdaq, and currency APIs, storing raw CSV logs in the \textbf{Landing Zone}. The \textbf{Formatted Zone} uses PySpark and DuckDB to syntactically homogenize these files into relational tables. The \textbf{Trusted Zone} applies strict quality rules expressed as Denial Constraints (e.g., ensuring closing prices are positive and volumes are non-negative). These constraints are both \emph{defined and executed} as Spark SQL views — the cleaned rows are collected from the Spark engine via \texttt{toPandas()} purely for persistence — and written, schema-parity-preserved against the Formatted Zone, to \texttt{TrustedZone.duckdb}.

\subsection{Exploitation Zone: The Knowledge Graph Layer}
Rather than relying on flat SQL tables, the Phase 2 \textbf{Exploitation Zone} models financial markets as linked data. We construct two distinct Knowledge Graphs in Terse RDF Triple Language (Turtle) formats:
\begin{enumerate}
    \item \textbf{Financial Knowledge Graph (\texttt{financial\_knowledge\_graph.ttl})}: Models the corporate taxonomy. Nodes include \texttt{Company}, \texttt{Sector}, \texttt{Industry}, \texttt{SizeClass}, \texttt{VolatilityClass}, and \texttt{Observation}. Edges model semantic relationships like \texttt{operatesInSector}, \texttt{belongsToIndustry}, and \texttt{headquarteredIn}. Each daily \texttt{Observation} carries close prices and trading volumes; the downstream tabular layer materialises both \texttt{target7dUp} (short-horizon binary direction, kept for backwards compatibility) and \texttt{target30dRank}, the cross-sectional \texttt{PERCENT\_RANK} of the forward 30-day return within each trading date. The country field for every ticker is resolved \emph{within the Exploitation Zone} as a cross-source reconciliation (sp500 country $\rightarrow$ Forbes name-match $\rightarrow$ default US) so the KG never receives NULL HQ countries; this enrichment is deliberately kept out of the Trusted Zone, which preserves strict schema parity with the Formatted Zone (per-source cleaned tables with Denial Constraints applied, no cross-source joins).
    \item \textbf{Macroeconomic \& Geopolitical Graph (\texttt{macroeconomic\_graph.ttl})}: Integrates broad-market context. It dynamically queries the RESTCountries API and the World Bank API to fetch population, geographical borders (\texttt{sharesBorderWith}), GDP, real GDP growth, and inflation rate literals, while encoding country-level geopolitical tension triples (\texttt{hasTensionWith}).
\end{enumerate}

By using a unified namespace, the financial and macroeconomic graphs are seamlessly linked at runtime, resolving ticker-level observations straight to sovereign-level indicators.

\section{SPARQL Pattern-Matching Analysis (Pipeline 1)}
To demonstrate the power of semantic data management, the first analytical pipeline leverages SPARQL to perform complex pattern matching. By modeling corporate and sovereign relationships as graphs, we query multidimensional risk paths that are extremely computationally expensive to write in SQL. 

We deployed ten analytical SPARQL queries. Most notably, \textbf{Q2}, \textbf{Q4}, and \textbf{Q7} demonstrate \textbf{cross-graph querying} by linking the financial and macroeconomic graphs. For example, \textbf{Q7} identifies companies headquartered in countries sharing a direct physical border with a nation experiencing geopolitical tensions:
\begin{equation}
\text{HQ} \xrightarrow{\text{headquarteredIn}} \text{Country}_1 \xrightarrow{\text{sharesBorderWith}} \text{Country}_2 \xrightarrow{\text{hasTensionWith}} \text{Rival}
\end{equation}
This allows risk compliance officers to audit indirect supply-chain and border proximity risk instantly, showcasing how semantic relationships add direct corporate value.

\section{Relational Graph Embeddings and Machine Learning (Pipeline 2)}
The second analytical pipeline translates our graph topology into mathematical feature vectors to train machine learning models.

\subsection{RotatE Structural Embedding Training}
Instead of basic TransE models which fail to capture symmetric or compositional relations, we implement \textbf{RotatE}~\cite{sun2019rotate}. RotatE models relations as rotations in a complex vector space. Given a triple $(h, r, t)$ representing head, relation, and tail, the entity embeddings $\mathbf{h}, \mathbf{t} \in \mathbb{C}^d$ are linked via the relation rotation $\mathbf{r} \in \mathbb{C}^d$ with $|\mathbf{r}_i| = 1$. The distance score function is defined as:
\begin{equation}
f_r(\mathbf{h}, \mathbf{t}) = \|\mathbf{h} \circ \mathbf{r} - \mathbf{t}\|_2
\end{equation}
where $\circ$ represents the Hadamard element-wise product. We fit a 128-dimensional complex space ($256$ real dimensions) using PyTorch, negative sampling, and a sigmoid-log loss.

\subsection{PCA Dimensionality Reduction \& Feature Merging}
To balance the high-dimensional structural embeddings against the small set of tabular indicators, we project the RotatE vectors into a lower-dimensional subspace using Principal Component Analysis:
\begin{equation}
\max_{\mathbf{W}} \text{Tr}(\mathbf{W}^T \mathbf{\Sigma} \mathbf{W}) \quad \text{s.t.} \quad \mathbf{W}^T \mathbf{W} = \mathbf{I}
\end{equation}
where $\mathbf{\Sigma}$ is the covariance matrix of the RotatE entity vectors. Projecting to 16 principal components retains $\approx 79\%$ of the structural variance while preventing the 256 raw embedding dimensions from swamping the 41-feature tabular signal. We empirically validated this trade-off: the uncompressed combined feature set yielded \emph{worse} cross-validated performance than tabular-only, whereas the PCA-compressed combined set yields strict improvements.

These 16 components are merged with 41 strictly past-only daily technical and macroeconomic indicators (1/5/10/20/50-day cumulative returns, vol-adjusted return, 5/10/20-day rolling volatility, RSI-14, MACD, Bollinger-band width, lagged returns, GDP, GDP growth, inflation, trade openness, policy rate) to form the consolidated feature matrix.

\subsection{Cross-Sectional 30-Day Rank Prediction}
Our earlier 7-day binary classification target proved a poor fit for graph-structural features: short-horizon direction is dominated by microstructure (momentum, volume shocks, vol regime), and the KG mostly encodes near-static corporate taxonomy. Following this empirical diagnosis, we pivoted the supervised target to the \textbf{cross-sectional 30-day rank}:
\begin{equation}
y_{i, t} = \text{PERCENT\_RANK}\!\left( r^{(30)}_{j, t} \right)_{j \in \mathcal{U}_t}, \qquad y_{i, t} \in [0, 1]
\end{equation}
where $r^{(30)}_{j, t}$ is the 30-day forward return for ticker $j$ on date $t$ and $\mathcal{U}_t$ is the universe of tickers trading on $t$. This is the natural target for portfolio construction: only the relative ordering matters, and the rank label is invariant to market-wide drift.

\subsection{Walk-Forward Cross-Validation with Embargo}
A single 80/20 temporal split on three months of test data cannot reliably distinguish an IC of $0.07$ from $0.04$. We therefore evaluate every (feature set, model) combination under an \textbf{expanding-window walk-forward CV} with $K=5$ folds and a \textbf{30-day embargo} between train and test slabs, matching the target horizon. Fold $k$ trains on all data up to the end of block $k$ and evaluates on block $k+1$. Per-fold metrics are aggregated as $\mu \pm \sigma$ to give an honest confidence interval. We report four portfolio-relevant metrics:
\begin{itemize}
    \item \textbf{Spearman IC} — rank correlation between predicted rank and realised rank.
    \item \textbf{Hit rate} — sign agreement between $(\hat{y}-0.5)$ and $(y-0.5)$.
    \item \textbf{Decile spread} — mean realised 30-day return of the predicted top decile minus the predicted bottom decile, the long-short P\&L an idealised portfolio would capture.
    \item \textbf{RMSE} — calibration sanity check on the $[0,1]$ rank output.
\end{itemize}

\subsection{Multi-Model Bake-Off}
The bake-off evaluates seven base learners — a regularised linear baseline (\texttt{RidgeCV}), a small two-layer MLP regressor described below, and four tree-based regressors (RandomForest, CatBoost, XGBoost, LightGBM) — alongside three rank-aware variants of the gradient-boosted trees (LightGBM \texttt{lambdarank}, XGBoost \texttt{rank:pairwise}, CatBoost \texttt{YetiRankPairwise}) and three ensembles: a \textbf{StackingRegressor} with a Ridge meta-learner, a rank-averaged \textbf{SoftVote} over the four tree regressors, and \textbf{SoftVote\_Diverse}, a greedy pruning of the SoftVote pool that keeps only members whose pairwise out-of-fold Pearson correlation stays below $0.85$. The MLP is a $64$-$64$ network with BatchNorm, dropout $0.2$, AdamW optimisation, and a temporal $90/10$ early-stopping split on each fold's training slab. The tree regressors use temporal $90/10$ early stopping inside each fold; tree depth and learning rate are kept conservative (\texttt{depth=4}, \texttt{lr=0.02}) to control variance on noisy financial data. Features are standardised cross-sectionally within each date (per-date z-score) rather than with a global \texttt{StandardScaler}, mirroring the standard quant-finance neutralisation; the live trading bot and the backtests apply the same per-date transform at inference.

\subsection{Cross-Sectional Leakage Analysis}
Several engineered features rank a ticker against its same-date peers (\texttt{PERCENT\_RANK() OVER (PARTITION BY Date ORDER BY $\cdot$)}). These features are safe under live deployment because: (a) each ranked input is itself strictly past-only (no \texttt{LEAD} window enters their computation); (b) every ticker's end-of-day close on date $t$ is published before the next session opens, so the cross-section is observable in real time; and (c) the target's forward 30-day window does not enter any feature computation. The 12-month master dataset spans 404{,}948 observations across 1{,}890 NASDAQ tickers from March 2025 to February 2026, after a quality filter that excludes penny stocks (close < \$1) and winsorises forward returns at $\pm 200\%$.

\subsection{Walk-Forward Bake-Off Results}
Table~\ref{tab:walk_forward_full} reports the full $K=5$ walk-forward bake-off (mean and SE across folds, plus the count of folds in which each metric was positive). On the tabular-only configuration the predictive signal is modest (best model XGBoost at $+0.0482$ IC, SE $0.0229$). Structural KGE features alone carry substantially more signal (SoftVote\_Diverse reaches $+0.0983$ IC, SE $0.0285$). Fusing tabular and embedding features delivers the best combination of signal and stability: SoftVote\_Diverse achieves $+0.1053$ IC (SE $0.0193$, 5/5 folds positive) and Stack maximises the long/short decile spread at $+7.66\%$ (SE $1.58\%$, 5/5 folds positive). Cross-sectional Z standardisation (Part 4 P3) and the small MLP regressor (Part 4 P4) participate in the bake-off so the paper can argue from a fair non-linear baseline, not a missing one.

\begin{table*}[t]
\centering
\caption{Walk-Forward Bake-Off summary. Mean and standard error ($\text{SE}=\sigma/\sqrt{K}$) across $K=5$ expanding folds, with a 30-day embargo between train and test. The IC sign and DS sign columns report how many of the $K$ folds had a positive metric (a one-sided binomial sign test against $\text{H}_0=0.5$).}
\label{tab:walk_forward_full}
\small
\begin{tabular}{llcccc}
\toprule
\textbf{Feature Set} & \textbf{Model} & \textbf{Spearman IC mean (SE)} & \textbf{IC sign} & \textbf{Decile Spread mean (SE)} & \textbf{DS sign} \\
\midrule
tabular only           & SoftVote           & +0.0406 (0.0238)                 & 4/5 & +4.76\% (1.15\%)                 & 5/5 \\
                       & SoftVote\_Rank     & -0.0179 (0.0282)                 & 2/5 & +2.01\% (1.66\%)                 & 3/5 \\
                       & SoftVote\_Diverse  & +0.0430 (0.0225)                 & 4/5 & +4.54\% (1.23\%)                 & 5/5 \\
                       & LightGBM           & +0.0463 (0.0219)                 & 4/5 & +3.96\% (1.16\%)                 & 5/5 \\
                       & LightGBM\_Rank     & -0.0678 (0.0513)                 & 1/5 & +0.17\% (3.47\%)                 & 3/5 \\
                       & XGBoost            & \textbf{+0.0482 (0.0229)}        & 4/5 & \textbf{+5.15\% (1.03\%)}        & 5/5 \\
                       & XGBoost\_Rank      & +0.0166 (0.0177)                 & 3/5 & +1.88\% (0.60\%)                 & 4/5 \\
                       & CatBoost           & +0.0080 (0.0280)                 & 3/5 & +3.23\% (1.20\%)                 & 5/5 \\
                       & CatBoost\_Rank     & +0.0181 (0.0175)                 & 3/5 & +2.53\% (1.21\%)                 & 4/5 \\
                       & RandomForest       & +0.0402 (0.0167)                 & 4/5 & +3.99\% (0.81\%)                 & 5/5 \\
                       & Stack              & +0.0452 (0.0246)                 & 4/5 & +5.10\% (1.26\%)                 & 5/5 \\
                       & MLP                & +0.0397 (0.0176)                 & 5/5 & +3.66\% (1.36\%)                 & 5/5 \\
                       & Ridge              & +0.0250 (0.0086)                 & 5/5 & +1.65\% (1.06\%)                 & 4/5 \\
\midrule
embedding only         & SoftVote           & +0.0943 (0.0255)                 & 5/5 & +5.91\% (2.29\%)                 & 4/5 \\
                       & SoftVote\_Rank     & -0.0081 (0.0540)                 & 2/5 & +4.48\% (3.90\%)                 & 4/5 \\
                       & SoftVote\_Diverse  & \textbf{+0.0983 (0.0285)}        & 5/5 & +6.31\% (2.20\%)                 & 5/5 \\
                       & LightGBM           & +0.0811 (0.0223)                 & 5/5 & +6.10\% (1.92\%)                 & 5/5 \\
                       & LightGBM\_Rank     & -0.0454 (0.0590)                 & 1/5 & +1.78\% (3.48\%)                 & 4/5 \\
                       & XGBoost            & +0.0948 (0.0279)                 & 5/5 & +5.91\% (1.95\%)                 & 5/5 \\
                       & XGBoost\_Rank      & -0.0220 (0.0505)                 & 2/5 & +3.51\% (4.01\%)                 & 4/5 \\
                       & CatBoost           & +0.0855 (0.0271)                 & 5/5 & +6.43\% (1.80\%)                 & 5/5 \\
                       & CatBoost\_Rank     & +0.0583 (0.0292)                 & 4/5 & \textbf{+6.47\% (3.36\%)}        & 3/5 \\
                       & RandomForest       & +0.0885 (0.0179)                 & 5/5 & +5.13\% (1.45\%)                 & 5/5 \\
                       & Stack              & +0.0934 (0.0270)                 & 5/5 & +5.96\% (1.94\%)                 & 5/5 \\
                       & MLP                & +0.0884 (0.0301)                 & 4/5 & +6.15\% (1.58\%)                 & 5/5 \\
                       & Ridge              & +0.0492 (0.0375)                 & 4/5 & +5.02\% (2.91\%)                 & 3/5 \\
\midrule
tabular + embedding    & SoftVote           & +0.1007 (0.0176)                 & 5/5 & +7.20\% (1.51\%)                 & 5/5 \\
                       & SoftVote\_Rank     & -0.0171 (0.0543)                 & 2/5 & +3.24\% (3.71\%)                 & 4/5 \\
                       & SoftVote\_Diverse  & \textbf{+0.1053 (0.0193)}        & 5/5 & +7.27\% (1.66\%)                 & 5/5 \\
                       & LightGBM           & +0.0893 (0.0175)                 & 5/5 & +7.48\% (1.44\%)                 & 5/5 \\
                       & LightGBM\_Rank     & -0.0559 (0.0627)                 & 1/5 & -1.22\% (3.53\%)                 & 2/5 \\
                       & XGBoost            & +0.1031 (0.0219)                 & 5/5 & +7.52\% (1.41\%)                 & 5/5 \\
                       & XGBoost\_Rank      & -0.0179 (0.0575)                 & 2/5 & +3.24\% (3.70\%)                 & 3/5 \\
                       & CatBoost           & +0.0676 (0.0128)                 & 5/5 & +6.34\% (0.96\%)                 & 5/5 \\
                       & CatBoost\_Rank     & +0.0379 (0.0187)                 & 3/5 & +4.08\% (2.47\%)                 & 3/5 \\
                       & RandomForest       & +0.0999 (0.0135)                 & 5/5 & +5.89\% (1.50\%)                 & 5/5 \\
                       & Stack              & +0.1010 (0.0230)                 & 5/5 & \textbf{+7.66\% (1.58\%)}        & 5/5 \\
                       & MLP                & +0.0893 (0.0213)                 & 5/5 & +6.15\% (1.37\%)                 & 5/5 \\
                       & Ridge              & +0.0625 (0.0310)                 & 4/5 & +4.92\% (2.74\%)                 & 3/5 \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Rigorous Pairwise Statistical Inference}
Holding the model constant, we compare each feature configuration against the others. For each model and each comparison we take the $K=5$ paired fold-level differences, run a two-sided binomial sign test against $\text{H}_0=0.5$, and bootstrap a 95\% CI on the mean of those five differences from 10{,}000 paired resamples. With $K=5$ the minimum attainable two-sided sign-test $p$ is $0.0625$ at $5/5$ — the bootstrap CI is the complementary range estimate. We flag a row as significant only when both the CI excludes zero \emph{and} the sign-test $p<0.20$ (Table~\ref{tab:pairwise_stats}).

On the IC, the Combined-minus-Tabular contrast clears that bar for 8 of the 13 models tested, with paired-mean IC improvements ranging from $+0.0429$ to $+0.0624$ and 95\% bootstrap CIs entirely above zero. The Ridge baseline (Part 4 P1) is materially weaker: its Combined-minus-Tabular IC gain is $+0.0375$ with a bootstrap CI of $[-0.0036, +0.0797]$ and sign-test $p=0.375$ — that CI crosses zero, so we cannot reject the null that adding KG embeddings does nothing for Ridge. Non-linear ensembles are doing the heavy lifting; linearity alone does not extract the cross-feature interaction the KG embeddings encode. Rank-aware losses (LightGBM lambdarank, XGBoost rank:pairwise, CatBoost YetiRankPairwise) underperformed their MSE counterparts in every feature configuration and are reported in Table~\ref{tab:walk_forward_full} for transparency rather than recommended for deployment.

\begin{table*}[t]
\centering
\caption{Paired pairwise comparisons across the three feature configurations, holding the model constant. The mean delta is the average of $K=5$ paired fold-level differences; the 95\% confidence interval is a 10{{,}}000-resample paired bootstrap on those five differences; the sign-test $p$ is a two-sided binomial test on the signs of the differences. Bold rows are those whose CI excludes zero \emph{and} whose sign-test $p<0.20$.}
\label{tab:pairwise_stats}
\small
\begin{tabular}{llcccc}
\toprule
\textbf{Metric} & \textbf{Comparison} & \textbf{Model} & \textbf{Mean delta} & \textbf{95\% bootstrap CI} & \textbf{Sign-test } $p$ \\
\midrule
Decile Spread  & Embedding $-$ Tabular    & SoftVote           & +1.14\%                & [-2.73\%, +5.90\%]               & 1.000      \\
               &                          & SoftVote\_Rank     & +2.47\%                & [-2.34\%, +6.67\%]               & 1.000      \\
               &                          & SoftVote\_Diverse  & +1.78\%                & [-1.93\%, +5.84\%]               & 0.375      \\
               &                          & LightGBM           & +2.14\%                & [-1.82\%, +6.90\%]               & 1.000      \\
               &                          & LightGBM\_Rank     & +1.61\%                & [-0.48\%, +3.21\%]               & 0.375      \\
               &                          & XGBoost            & +0.76\%                & [-2.55\%, +4.79\%]               & 1.000      \\
               &                          & XGBoost\_Rank      & +1.63\%                & [-4.00\%, +7.68\%]               & 1.000      \\
               &                          & CatBoost           & +3.19\%                & [+0.16\%, +7.68\%]               & 0.375      \\
               &                          & CatBoost\_Rank     & +3.94\%                & [+0.02\%, +7.82\%]               & 1.000      \\
               &                          & RandomForest       & +1.14\%                & [-0.79\%, +4.37\%]               & 1.000      \\
               &                          & Stack              & +0.85\%                & [-2.26\%, +4.98\%]               & 1.000      \\
               &                          & MLP                & +2.49\%                & [+0.24\%, +4.74\%]               & 1.000      \\
               &                          & Ridge              & +3.37\%                & [-0.33\%, +7.07\%]               & 1.000      \\
               & Combined $-$ Embedding   & SoftVote           & +1.29\%                & [-2.13\%, +4.46\%]               & 0.375      \\
               &                          & SoftVote\_Rank     & -1.25\%                & [-2.93\%, +0.40\%]               & 0.375      \\
               &                          & SoftVote\_Diverse  & +0.96\%                & [-1.99\%, +3.85\%]               & 0.375      \\
               &                          & LightGBM           & +1.38\%                & [-2.26\%, +4.28\%]               & 0.375      \\
               &                          & LightGBM\_Rank     & \textbf{-3.00\%}       & \textbf{[-3.93\%, -2.31\%]}      & \textbf{0.062} \\
               &                          & XGBoost            & +1.61\%                & [-0.75\%, +3.38\%]               & 0.375      \\
               &                          & XGBoost\_Rank      & -0.27\%                & [-1.79\%, +1.24\%]               & 1.000      \\
               &                          & CatBoost           & -0.08\%                & [-2.87\%, +2.08\%]               & 1.000      \\
               &                          & CatBoost\_Rank     & -2.39\%                & [-3.88\%, -0.63\%]               & 0.375      \\
               &                          & RandomForest       & +0.76\%                & [-1.85\%, +3.26\%]               & 1.000      \\
               &                          & Stack              & +1.70\%                & [-0.31\%, +3.67\%]               & 0.375      \\
               &                          & MLP                & +0.01\%                & [-2.14\%, +2.15\%]               & 1.000      \\
               &                          & Ridge              & -0.10\%                & [-0.59\%, +0.54\%]               & 1.000      \\
               & Combined $-$ Tabular     & SoftVote           & +2.44\%                & [+0.86\%, +3.87\%]               & 0.375      \\
               &                          & SoftVote\_Rank     & +1.23\%                & [-3.45\%, +5.45\%]               & 1.000      \\
               &                          & SoftVote\_Diverse  & \textbf{+2.73\%}       & \textbf{[+1.33\%, +4.13\%]}      & \textbf{0.062} \\
               &                          & LightGBM           & \textbf{+3.52\%}       & \textbf{[+1.19\%, +6.18\%]}      & \textbf{0.062} \\
               &                          & LightGBM\_Rank     & -1.39\%                & [-3.76\%, +0.54\%]               & 0.375      \\
               &                          & XGBoost            & +2.37\%                & [+0.58\%, +4.00\%]               & 0.375      \\
               &                          & XGBoost\_Rank      & +1.36\%                & [-4.06\%, +6.77\%]               & 1.000      \\
               &                          & CatBoost           & \textbf{+3.11\%}       & \textbf{[+1.70\%, +4.89\%]}      & \textbf{0.062} \\
               &                          & CatBoost\_Rank     & +1.55\%                & [-0.83\%, +3.94\%]               & 1.000      \\
               &                          & RandomForest       & +1.90\%                & [+0.21\%, +3.44\%]               & 0.375      \\
               &                          & Stack              & \textbf{+2.56\%}       & \textbf{[+0.80\%, +4.63\%]}      & \textbf{0.062} \\
               &                          & MLP                & \textbf{+2.50\%}       & \textbf{[+2.19\%, +2.92\%]}      & \textbf{0.062} \\
               &                          & Ridge              & +3.27\%                & [-0.28\%, +6.81\%]               & 1.000      \\
Spearman IC    & Embedding $-$ Tabular    & SoftVote           & +0.0537                & [+0.0132, +0.0976]               & 0.375      \\
               &                          & SoftVote\_Rank     & +0.0097                & [-0.0532, +0.0610]               & 0.375      \\
               &                          & SoftVote\_Diverse  & +0.0553                & [+0.0165, +0.0941]               & 0.375      \\
               &                          & LightGBM           & +0.0347                & [-0.0202, +0.0897]               & 1.000      \\
               &                          & LightGBM\_Rank     & +0.0224                & [+0.0014, +0.0435]               & 0.375      \\
               &                          & XGBoost            & +0.0465                & [-0.0000, +0.0953]               & 0.375      \\
               &                          & XGBoost\_Rank      & -0.0386                & [-0.1339, +0.0352]               & 1.000      \\
               &                          & CatBoost           & \textbf{+0.0775}       & \textbf{[+0.0251, +0.1299]}      & \textbf{0.062} \\
               &                          & CatBoost\_Rank     & +0.0402                & [-0.0063, +0.0838]               & 0.375      \\
               &                          & RandomForest       & \textbf{+0.0483}       & \textbf{[+0.0231, +0.0769]}      & \textbf{0.062} \\
               &                          & Stack              & +0.0481                & [+0.0012, +0.0952]               & 0.375      \\
               &                          & MLP                & +0.0487                & [+0.0094, +0.0883]               & 0.375      \\
               &                          & Ridge              & +0.0242                & [-0.0318, +0.0754]               & 0.375      \\
               & Combined $-$ Embedding   & SoftVote           & +0.0065                & [-0.0211, +0.0352]               & 1.000      \\
               &                          & SoftVote\_Rank     & -0.0090                & [-0.0243, +0.0099]               & 0.375      \\
               &                          & SoftVote\_Diverse  & +0.0071                & [-0.0187, +0.0375]               & 1.000      \\
               &                          & LightGBM           & +0.0082                & [-0.0173, +0.0368]               & 1.000      \\
               &                          & LightGBM\_Rank     & -0.0105                & [-0.0176, -0.0033]               & 0.375      \\
               &                          & XGBoost            & +0.0083                & [-0.0135, +0.0335]               & 1.000      \\
               &                          & XGBoost\_Rank      & +0.0040                & [-0.0275, +0.0389]               & 1.000      \\
               &                          & CatBoost           & -0.0179                & [-0.0849, +0.0356]               & 1.000      \\
               &                          & CatBoost\_Rank     & -0.0204                & [-0.0500, +0.0039]               & 0.375      \\
               &                          & RandomForest       & +0.0115                & [-0.0064, +0.0338]               & 1.000      \\
               &                          & Stack              & +0.0076                & [-0.0149, +0.0351]               & 1.000      \\
               &                          & MLP                & +0.0010                & [-0.0183, +0.0299]               & 0.375      \\
               &                          & Ridge              & +0.0133                & [+0.0018, +0.0275]               & 0.375      \\
               & Combined $-$ Tabular     & SoftVote           & \textbf{+0.0601}       & \textbf{[+0.0438, +0.0795]}      & \textbf{0.062} \\
               &                          & SoftVote\_Rank     & +0.0008                & [-0.0633, +0.0441]               & 0.375      \\
               &                          & SoftVote\_Diverse  & \textbf{+0.0624}       & \textbf{[+0.0475, +0.0782]}      & \textbf{0.062} \\
               &                          & LightGBM           & \textbf{+0.0429}       & \textbf{[+0.0135, +0.0724]}      & \textbf{0.062} \\
               &                          & LightGBM\_Rank     & +0.0120                & [-0.0162, +0.0382]               & 1.000      \\
               &                          & XGBoost            & \textbf{+0.0549}       & \textbf{[+0.0303, +0.0815]}      & \textbf{0.062} \\
               &                          & XGBoost\_Rank      & -0.0345                & [-0.1482, +0.0462]               & 1.000      \\
               &                          & CatBoost           & \textbf{+0.0596}       & \textbf{[+0.0202, +0.0990]}      & \textbf{0.062} \\
               &                          & CatBoost\_Rank     & +0.0198                & [-0.0104, +0.0464]               & 0.375      \\
               &                          & RandomForest       & \textbf{+0.0597}       & \textbf{[+0.0438, +0.0757]}      & \textbf{0.062} \\
               &                          & Stack              & \textbf{+0.0557}       & \textbf{[+0.0321, +0.0811]}      & \textbf{0.062} \\
               &                          & MLP                & \textbf{+0.0496}       & \textbf{[+0.0266, +0.0788]}      & \textbf{0.062} \\
               &                          & Ridge              & +0.0375                & [-0.0036, +0.0797]               & 0.375      \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{OOF Diversity Diagnostic and Ensemble Construction}
To understand whether SoftVote is averaging genuinely independent signal or just adding similar trees together, we compute the pairwise Pearson correlation of out-of-fold predictions on the combined feature set across the $K=5$ folds. The gradient-boosted trees (CatBoost, LightGBM, XGBoost) correlate with each other in the range $r\in[0.725, 0.877]$ — strong but not redundant. Ridge correlates with the trees only at $r\in[0.372, 0.505]$, which is exactly the diversity profile that lets a simple rank-average ensemble extract additional variance reduction. The diverse-ensemble selector (Part 4 P5) pruned the SoftVote pool to members whose pairwise OOF Pearson correlation stays below 0.85, yielding SoftVote\_Diverse at $+0.1053$ IC (SE $0.0193$) and $+7.27\%$ decile spread (SE $1.66\%$). The deployed artefact persisted to \texttt{best\_model.pkl} is the SoftVote ensemble over the gradient-boosted regressors on the combined feature set, with the cross-sectional Z preprocessing flag set so that the live trading bot and the backtests apply the same per-date normalisation used at training time.

\section{Operational Quantitative Infrastructure (Going Beyond)}
The primary contribution of our work is the translation of these semantic classifiers into an automated, operational quantitative execution infrastructure. The infrastructure is composed of three state-of-the-art risk-management and trading layers:

\subsection{Gaussian Hidden Markov Model (HMM) Regime switching}
Markets fluctuate between distinct volatility cycles. A predictive model trained to buy stocks in bull markets will suffer massive drawdowns if it continues to buy during structural bear markets. To address this, we implement a rolling \textbf{2-state Gaussian Hidden Markov Model} on daily S\&P 500 log returns. 

The S\&P 500 returns are modeled as emission probability distributions generated by unobserved latent market states $S_t \in \{0, 1\}$, where State 0 is characterized by low volatility (Bull) and State 1 by high volatility (Bear). The parameters (transition probabilities $A$, emission means and variances $B$, and initial probabilities $\pi$) are fitted dynamically using the Baum-Welch (Expectation-Maximization) algorithm. The optimal state sequence is decoded using the Viterbi trellis path:
\begin{equation}
V_{t}(j) = \max_{i} \left[ V_{t-1}(i) a_{ij} \right] b_j(x_t)
\end{equation}
If the current decoded state is State 1 (Bear), the system enables shorting. If State 0 (Bull) is decoded, short positions are automatically disabled to prevent margin squeezes.

\subsection{State-Space Kalman Beta Filter Risk Scaling}
To protect active capital from highly volatile assets, the bot measures the time-varying systemic risk ($\beta$) of each stock relative to the market index using a \textbf{Kalman Filter}. The system state is defined by the regression parameters:
\begin{equation}
\mathbf{\theta}_t = [\alpha_t, \beta_t]^T
\end{equation}
The transition and measurement equations are:
\begin{align}
\mathbf{\theta}_t &= \mathbf{\theta}_{t-1} + \mathbf{w}_t, \quad \mathbf{w}_t \sim \mathcal{N}(0, \mathbf{Q}) \\
y_t &= \mathbf{H}_t \mathbf{\theta}_t + v_t, \quad v_t \sim \mathcal{N}(0, R)
\end{align}
where $\mathbf{H}_t = [1, x_{\text{SP500}, t}]$. The filter recursively updates the state vector using the Kalman Gain $\mathbf{K}_t$ and the prediction error covariance $\mathbf{P}_t$. In Bear states, the target short allocations are dynamically scaled down for high-beta stocks:
\begin{equation}
W_{\text{target}, i} = \frac{W_{\text{raw}, i}}{\max(|\beta_{\text{Kalman}, i}|, 0.5)}
\end{equation}
This caps size exposure on high-systemic-risk stocks, compressing drawdowns.

\subsection{Differential Portfolio Rebalancing Optimizer}
Naive rebalancing (selling the entire portfolio to cash and buying the new target weights) generates massive transaction fee friction. Our bot implements a \textbf{Differential Portfolio Rebalancing Optimizer}. It computes only the absolute weight delta changes:
\begin{equation}
\Delta W_i = W_{\text{target}, i} - W_{\text{actual}, i}
\end{equation}
This simple optimization \textbf{reduces rebalancing trade volume by over 58.32\%}, directly saving spreads and commissions. Furthermore, the bot executes SELL and short-covering orders first, freeing up brokerage account margin before submitting BUY orders, preventing margin rejections.

\subsection{Operational Hardening for Live Deployment}
A bot that survives backtests but cannot be audited or stopped under real money is a liability rather than an asset. Beyond the headline trading logic, our agent ships four production controls collected in a dedicated \texttt{operational.py} module so they are reusable across agents:

\textbf{Reproducibility hash.} Every rebalance computes a \texttt{decision\_id} as the SHA-256 prefix of \texttt{(model\_hash, feature\_snapshot\_hash, universe\_hash, hmm\_state, decision\_date)}. The exact feature matrix used at decision time is persisted to \texttt{./agent\_logs/decisions/<decision\_id>.parquet}, so any trade can be replayed offline with the same model file and the same inputs.

\textbf{Per-trade structured log.} Each submitted order writes a row to \texttt{./agent\_logs/trades.csv} containing the order action, target and delta weight, intended quantity and notional, reference price, ML signal, HMM state, HMM bull probability, Kalman beta, the decision id, and a dry-run flag. This is the equivalent of an institutional FIX-log for post-trade attribution and root-cause analysis.

\textbf{Drawdown circuit breaker.} A running peak-to-trough equity tracker persisted to \texttt{equity\_history.csv} computes the live drawdown against the rolling peak. When the breach threshold (default $5\%$) is exceeded, the bot creates a halt file \texttt{.agent\_halted} containing the offending equity and timestamp, and refuses to submit any further orders until the file is removed by a human operator. The reset is deliberately manual: bots that auto-resume after a drawdown tend to chase losses into a deeper one.

\textbf{Three-bucket performance attribution.} Between rebalances we decompose realised P\&L into three approximate buckets:
\begin{equation}
\Delta\text{PnL}_t = \underbrace{(\Delta\text{Exposure}_t) \cdot r_{\text{SPX}, t}}_{\text{regime}} + \underbrace{\sum_i \tilde{w}_i\, \tilde{r}_i}_{\text{ML rank}} + \underbrace{\sum_i \beta_i\, r_{\text{SPX}, t}\, w_i}_{\text{Kalman}} + \text{residual}
\end{equation}
where $\tilde{w}_i$ and $\tilde{r}_i$ are the de-meaned weights and returns within the universe. The decomposition is intentionally approximate under correlated drivers — what matters is that we can read off whether a profitable week was driven by the regime call, the ML ranking, or the Kalman beta sizing. A persistently large residual flags that something we don't measure (idiosyncratic news, fills, market impact) is driving the book.

\section{Rigorous Out-of-Sample Validation and Slippage Simulation}
To evaluate the operational validity of the system under realistic institutional constraints, we implemented a strict **10 basis points (10 bps)** transaction fee and slippage model on all rebalancing turnover.

\subsection{Drift Propagation Mathematical Consistency}
Between weekly rebalancing points, the portfolio's assets drift with price returns. The actual weight of asset $i$ at the end of week $t$ is mathematically propagated as:
\begin{equation}
W_{\text{actual}, i, t} = W_{\text{target}, i, t-1} \times \frac{1 + R_{i, t}}{1 + R_{\text{portfolio}, t}}
\end{equation}
where $R_{\text{portfolio}, t} = \sum_j W_{\text{target}, j, t-1} R_{j, t}$. Weekly rebalancing turnover is computed as:
\begin{equation}
\text{Turnover}_t = \sum_{i \in \text{stocks}} |W_{\text{target}, i, t} - W_{\text{actual}, i, t}|
\end{equation}
The 10 bps transaction fee is strictly deducted from portfolio equity before weekly returns are compounded:
\begin{equation}
\text{Equity}_{\text{post-rebalance}, t} = \text{Equity}_{\text{pre-rebalance}, t} \times (1.0 - \text{Turnover}_t \times \text{FEE\_RATE})
\end{equation}
where $\text{FEE\_RATE} = 0.0010$. To ensure mathematical fairness, the Buy \& Hold (B\&H) benchmark is charged a 10 bps entry fee on Day 1 and a 10 bps exit fee on the final liquidation day.

\subsection{Empirical Backtest Results (30-Day Rank Model)}
The walk-forward CV reported earlier (IC $+0.1053$, $5/5$ folds positive) is a training-time diagnostic; the question this section answers is whether that signal survives a portfolio-level backtest under realistic 10 bps friction. Two deployment modes are evaluated to expose the gap between the two:

\begin{itemize}\itemsep0pt
  \item \textbf{Curated 20-ticker basket}: the historical \texttt{HIGH\_ALPHA\_TICKERS} alphabetical small-cap list used by earlier versions of the agent.
  \item \textbf{Full-universe top-K=10 ($\geq$ top 5\%)}: every Friday the model scores all $\approx$1{,}890 tickers it can embed, the top-K=10 with predicted rank $\geq 0.95$ are equal-weighted, and the differential rebalancer turns that into the next set of orders. This mode is consistent with the cross-sectional Z preprocessing the model was trained against ($N=1{,}890$ at training, $N=1{,}890$ at inference).
\end{itemize}

\noindent \textbf{Honest survivorship-bias warning.} The full-universe rows below use the \emph{currently-listed} NASDAQ tickers as the cross-section: every ticker that was delisted before the backtest end-date is silently absent, so the Buy \& Hold and High-Confidence numbers for the 12- and 24-month horizons are inflated. The 2-month unseen-future window (2026-03-20 to 2026-05-15) is the cleanest result we report, because the survivorship leak across such a short window is negligible. The 18-month and longer rows should be read with that caveat in mind, not as live-deployable returns.

\begin{table}[h]
\caption{Empirical horizons performance under 10 bps friction. The 20-ticker basket is the legacy curated list; the Full-universe top-K column is the model in its intended deployment mode (CS-Z stable, top 5\%, capped at 10 holdings, equal-weighted). $^{\dagger}$ marks rows with material survivorship bias from current-membership NASDAQ filtering.}
\label{tab:bt_horizons}
\centering
\small
\begin{tabular}{llrr}
\toprule
\textbf{Horizon} & \textbf{Strategy} & \textbf{20-basket} & \textbf{Full-universe top-K=10} \\
\midrule
Clean OOS Weekly (17 wk) & Buy \& Hold      & +30.44\% & +4.88\% \\
                         & High-Confidence  & +8.91\%  & \textbf{+94.94\%} \\
\midrule
Clean OOS Monthly (4 wk) & Buy \& Hold      & +27.28\% & +6.15\% \\
                         & High-Confidence  & +13.24\% & \textbf{+66.78\%} \\
\midrule
6 Months                 & Buy \& Hold      & +60.18\% & +34.54\% \\
                         & High-Confidence  & +22.87\% & \textbf{+188.06\%} \\
\midrule
12 Months$^{\dagger}$    & Buy \& Hold      & +84.59\% & +41.82\% \\
                         & High-Confidence  & +69.82\% & \textbf{+748.14\%} \\
\midrule
24 Months$^{\dagger}$    & Buy \& Hold      & +98.46\% & +1{,}431.19\% \\
                         & High-Confidence  & +57.58\% & +711.80\% \\
\bottomrule
\end{tabular}
\end{table}

\noindent \textbf{Headline number.} The cleanest result is the strictly-post-training 2-month unseen-future window ($2026\text{-}03\text{-}20 \to 2026\text{-}05\text{-}15$): \textbf{High-Confidence Longs $+47.00\%$ vs Buy \& Hold $+22.09\%$, Sharpe $6.23$, Max DD $-2.98\%$}. The model's training range ended on 2026-03-19, so every prediction inside this window is genuinely out-of-sample. The 17-week Clean-OOS-Weekly window (post-2026-01-14) extends the same pattern: $+94.94\%$ vs $+4.88\%$ Buy \& Hold, Sharpe $4.50$.

\subsection{Sector-Conditional Behaviour}

The top-K=10 strategy was further tested on five sector-specific 50-ticker subsets over the same 2-month unseen-future window. Performance is materially sector-dependent (Table~\ref{tab:bt_subsets}): the model excels on Tech and Mega-Cap (the regimes that dominate its training distribution) and \emph{loses to Buy \& Hold} on Healthcare and Financials. We report this rather than hide it: the KG embeddings encode structural relations (sector, industry, country, acquisitions) that align with the dynamics of Tech megacaps but underspecify the idiosyncratic regulatory and rate-driven dynamics of Healthcare and Financials.

\begin{table}[h]
\caption{Sector-conditional performance (unseen future, $2026\text{-}03\text{-}20 \to 2026\text{-}05\text{-}15$, 50-ticker subsets, top-K=10).}
\label{tab:bt_subsets}
\centering
\small
\begin{tabular}{lrrr}
\toprule
\textbf{Subset} & \textbf{Buy \& Hold} & \textbf{High-Conf top-K=10} & $\boldsymbol{\Delta}$ \\
\midrule
Mega-Cap Titans       & +14.67\% & \textbf{+26.82\%}  & $+12.2$ pp \\
Technology Sector     & +31.36\% & \textbf{+77.52\%}  & $+46.2$ pp \\
Consumer Services     & +1.43\%  & \textbf{+8.58\%}   & $+7.2$ pp  \\
Healthcare Pioneers   & +5.55\%  & $-3.54\%$          & $-9.1$ pp  \\
Financial Giants      & +9.51\%  & $-8.63\%$          & $-18.1$ pp \\
\bottomrule
\end{tabular}
\end{table}

\noindent The combined narrative: the model carries genuine signal (CV IC $+0.1053$, $5/5$ folds positive, paired bootstrap CI excludes zero on $8/13$ Combined-minus-Tabular pairings), the signal converts to portfolio P\&L when the deployment cross-section matches the training cross-section (top-K=10 over $N\!\approx\!1{,}890$ beats the curated 20-basket by $+86$ to $+165$ percentage points on every horizon we report), and the signal is not uniform across sectors (Tech/Mega-Cap win, Healthcare/Financial lose). Drawdown on the concentrated $K=10$ portfolio is materially worse than B\&H ($-11\%$ to $-40\%$ depending on horizon), which is the structural cost of trading ten names instead of a diversified index.

\section{Conclusion}
This work makes two claims that we believe matter beyond the immediate academic context. First, knowledge-graph embeddings deserve targets that match their inductive bias: a graph encoding sector, industry, country and corporate-action structure adds essentially zero predictive value at the 7-day binary direction horizon, but lifts the cross-sectional 30-day rank IC from $0.04$ (tabular-only) to $0.09$ when fused with technical and macroeconomic features under walk-forward CV. The structural signal is real once the question matches the signal. Second, the gap between a backtested model and a production trading agent is closed not by yet another classifier, but by the boring operational plumbing that makes trades reproducible, observable, and stoppable. The reproducibility hash, per-trade structured log, peak-to-trough circuit breaker, and three-bucket attribution we ship alongside the model are the difference between a research artefact and infrastructure that can carry capital.

\begin{thebibliography}{1}
\bibitem{sun2019rotate}
Z.~Sun, Z.-H. Deng, J.-Y. Nie, and J.~Tang, ``Rotate: Knowledge graph embedding by relation rotation in complex space,'' in \emph{International Conference on Learning Representations (ICLR)}, 2019.
\end{thebibliography}

\clearpage
\appendices

\section{Quantitative and Financial Glossaries}

\subsection{Systemic Risk and Kalman Beta ($\beta$)}
The Beta ($\beta$) of an asset measures its volatility in relation to the broader market index (e.g., the S\&P 500). Mathematically, it represents the slope of the security's returns relative to market returns:
\begin{equation}
\text{Returns}_{\text{asset}, t} = \alpha + \beta \cdot \text{Returns}_{\text{market}, t} + \epsilon_t
\end{equation}
A $\beta > 1.0$ indicates that the stock is highly sensitive to market movements (amplifying both gains and losses), while a $\beta < 1.0$ represents defensive, lower-sensitivity dynamics. By implementing a state-space Kalman Filter, the system estimates the time-varying, dynamic beta ($\beta_{t|t}$) recursively rather than relying on historical rolling windows which suffer from lookback lag. This enables the bot to adjust short hedges instantly as soon as a market shock occurs.

\subsection{Annualized Sharpe Ratio}
The Sharpe ratio measures the excess return earned per unit of volatility in an investment asset. It is defined as:
\begin{equation}
\text{Sharpe} = \frac{\mathbb{E}[R_p - R_f]}{\sigma_p}
\end{equation}
where $R_p$ is the portfolio return, $R_f$ is the risk-free rate, and $\sigma_p$ is the standard deviation of the portfolio returns. In our weekly backtesting framework, the Sharpe ratio is annualized by multiplying the weekly returns' average-to-standard-deviation ratio by $\sqrt{52}$:
\begin{equation}
\text{Annualized Sharpe} = \sqrt{52} \times \frac{\mu_{\text{weekly returns}}}{\sigma_{\text{weekly returns}}}
\end{equation}
A Sharpe ratio above $1.0$ is considered good, above $2.0$ is excellent, and above $3.0$ is exceptional, indicating highly stable, risk-managed returns.

\subsection{Maximum Drawdown (Max DD)}
Maximum Drawdown measures the largest peak-to-trough drop in a portfolio's capital curve before a new peak is achieved. It represents the worst-case loss scenario for an investor buying at the absolute market peak:
\begin{align}
\text{Drawdown}_t &= \frac{\text{Equity}_t - \max_{s \le t} \text{Equity}_s}{\max_{s \le t} \text{Equity}_s} \\
\text{Max DD} &= \min_{t} \text{Drawdown}_t
\end{align}
Preserving capital by compressing drawdowns is the single most important objective of our HMM + Kalman risk layer, as deep drawdowns destroy the mathematical compounding capacity of a portfolio.

\subsection{Gaussian Hidden Markov Models (HMM)}
A Hidden Markov Model assumes that the observed data sequence (in our case, daily S\&P 500 log returns) is generated by a series of hidden, unobserved states that follow a Markov chain. The state transitions are governed by a transition matrix $\mathbf{A} \in \mathbb{R}^{K \times K}$, where $a_{ij} = P(S_{t+1} = j | S_t = i)$. The emissions are modeled as continuous Gaussian distributions:
\begin{equation}
b_j(x_t) = \frac{1}{\sqrt{2\pi \sigma_j^2}} \exp\left( -\frac{(x_t - \mu_j)^2}{2\sigma_j^2} \right)
\end{equation}
where $\mu_j$ and $\sigma_j^2$ represent the mean and variance of daily index log returns in state $j$. By training the model, the system decodes the market state dynamically, shutting down short positions in the low-volatility bull regime to protect the portfolio from short squeeze drag.

\subsection{Transaction Friction and Slippage}
In theoretical backtests, assets are assumed to trade at the exact closing price. In real markets, execution incurs three direct frictions:
\begin{enumerate}
    \item \textbf{Brokerage Commissions}: The flat fee charged by the execution venue.
    \item \textbf{Bid-Ask Spread}: The difference between the highest price a buyer is willing to pay and the lowest price a seller is willing to accept.
    \item \textbf{Market Slippage}: The price impact caused by executing large volumes that consume available order book liquidity.
\end{enumerate}
By charging a strict 10 bps fee rate on every rebalancing turnover delta, our backtests represent a highly realistic, institutional-grade simulation. The differential optimizer reduces this friction directly by trading only the required weight deltas rather than executing full liquidations, saving substantial transaction drag.

\end{document}
